
\section{State of the Art}%
\label{sec:related_work}

This section examines the state of the art in tools and functionalities that support developement and interaction with Semantic Web technologies. 
Current implementations address various aspects of Semantic Web technologies, including for example, ontology modeling tools (e.g., Protégé) and SPARQL query editors (e.g., YASGUI).
Next we describe a set of common IDE functionalities~\cite{HowAreJava} and discuss existing tools that (partially) cover them for Semantic Web-related formats.


\subsection{Language Server Protocol}

The \textit{Language Server Protocol} (LSP), introduced by Microsoft, is designed primarily to facilitate reuse across different development environments.
It defines a \textit{JSON-RPC}-based communication protocol that allows a single language server implementation to be utilized across multiple editors.

Prior to the introduction of LSP, each combination of a programming language and an editor required a custom integration, leading to a complexity of \(\mathcal{O}(M \cdot N)\) for \(M\) editors and \(N\) languages\cite{Kj_r_Rask_2021}.
By decoupling language-specific analysis from editor-specific implementations, LSP reduces this complexity to \(\mathcal{O}(M + N)\), as only one editor-side implementation per editor and one language server per language are required.

In the following discussion of LSP features, we reference the corresponding \textit{JSON-RPC} requests where applicable.
Additionally, we highlight that not all LSP requests are implemented in the \textit{Semantic Web Language Server} (SWLS).
However, full implementation of all requests is not a strict requirement of the protocol, and is considered future work.


\subsubsection{Highlighting} (\texttt{textDocument/semanticTokens/full}) enhances readability by leveraging visual cues, which assist developers in understanding complex data.
For example by colouring differently reserved keywords (e.g., \texttt{a}, \texttt{@prefix} in the RDF Turtle syntax) and data types (e.g., strings, numbers, etc.).  
% Semantic highlighters should avoid overloading the user with unnecessary information. 
% Undefined properties, for example, should not be highlighted differently; such cases are better addressed by diagnostic features.~\todoCite{This needs to be backed by a citation. Otherwise just limit to explain what is highlighting and why is important.}
There are two primary types of highlighting: \textbf{syntactic} and \textbf{semantic}.

\begin{itemize}
    \item \textbf{Syntactic highlighting} focuses on document syntax and is implemented using lexers (i.e. software that transforms text into tokens: keyswords, IRIs, blanknode identifiers, etc.).
      For example in RDF Turtle, strings might appear in green, while named nodes are highlighted in red.
      This type of highlighting does not require knowledge of the document's meaning.
      Syntactic highlighting is enabled on all software we considered, except for the 
      \textit{rdf-vocabularies-autocomplete} VSCode extension, VSCode handles syntactic highlighting already with the Tree-sitter project\footnote{\url{https://tree-sitter.github.io/tree-sitter/}}.
    \item \textbf{Semantic highlighting}, (introduced by D. Nodlen\cite{DNolden}) in contrast, uses the local (and externally linked) semantics referenced in the document to differentiate terms based on their roles. 
      For instance, terms declared in an ontology can be highlighted differently compared to locally defined terms.
      We found no semantic web tooling that implemented this feature.
\end{itemize}


\subsubsection{Diagnostics} (\texttt{textDocument/diagnostic}) are essential for identifying issues early in the development process. 
Rather than deferring error detection to runtime, an IDE should provide immediate feedback on data validity and correctness.
These features enable developers to maintain data accuracy and reduce time spent debugging. 
The stardog language servers and the Yasgui query editor notify the users of syntax errors.

% Examples of valuable diagnostics include:
\begin{itemize}
    \item \textbf{Standard compliance}: Notifying users if the current document is incompatible with parsers adhering to Semantic Web standards.
    \item \textbf{Undefined properties}: Alerting users to potential typing mistakes or incorrect terms when undefined properties are detected.
    \item \textbf{Reasoning errors}: Highlighting logical inconsistencies in ontologies or data that deviate from expected SHACL shapes or OWL reasoning constraints.
\end{itemize}


\subsubsection{Formatting} (\texttt{textDocument/formatting})
Consistent formatting across documents can help enhancing readability and maintainability.
IDEs assist users in adhering to a defined style guide while allowing customization to accommodate different workflows or preferences.
We found no semantic web tooling that implemented this feature.

\subsubsection{Renaming} (\texttt{textDocument/rename}) identifiers, such as classes or properties, is an important feature for iterative development.
This functionality helps users manage data more efficiently and adapt to evolving project requirements without introducing errors.
We found no semantic web tooling except for our previous work~\cite{JSONLD-LSP}, that handles renaming identifiers.

\subsubsection{Completion} (\texttt{textDocument/completion}) features, similar to highlighting, can be categorized as \textbf{syntactic} or \textbf{semantic}:
Simple semantic completion might suggest all possible terms, but a sophisticated system should filter suggestions based on the context, such as the inferred type of the current subject or object.
All semantic web tooling we considered, included some kind of completion mechanism, none however applied typed completion.


\begin{itemize}
    \item \textbf{Syntactic completion} involves suggesting keywords or tokens already defined in the document.
    \item \textbf{Semantic completion} provides context-aware suggestions, such as properties or classes relevant to the current position. For example:
    \begin{itemize}
        \item When writing a property in an RDF document, the IDE can suggest those that define the subject's type as their \texttt{rdfs:domain}.
        \item When completing an object, it can suggest entities consistent with the property's \texttt{rdfs:range}.
    \end{itemize}
\end{itemize}

Table~\ref{tab:current_implementations} provides an overview of existing open-source tools that implement at least one of the IDE functionalities discussed in this section.
These tools focus on specific aspects, such as query editing, ontology modeling, or vocabulary completion, but none of them integrate a full set of IDE-like capabilities for Semantic Web technologies.

% we might want to transpose the headers
\begin{table}[h!]
    \centering
  \begin{tabularx}{\textwidth}{ |p{3cm}|C|C|C|C|C|C|C|C|C|C|C|C|}
\hline
    \multirow{2}{*}{} & \multicolumn{2}{c|}{Highlight} & \multicolumn{3}{c|}{Diagnostics} & \multicolumn{4}{c|}{Completion}  & \multirow{2}{*}{} & \multirow{2}{*}{} & \multirow{2}{*}{} \\ \cline{2-10} 

      LSP implementation                & \rotatebox{90}{Syntactic} & \rotatebox{90}{Semantic} %highlighting
                                        & \rotatebox{90}{Syntax} & \rotatebox{90}{Undefined} & \rotatebox{90}{Validation\ \ }  %diagnostics
                                        & \rotatebox{90}{Syntax} & \rotatebox{90}{Tokens} & \rotatebox{90}{Simple} & \rotatebox{90}{Typed} % completion
                                        & \rotatebox{90}{Formatting} 
                                        & \rotatebox{90}{Renaming}
                                        & \rotatebox{90}{Polyglot} \\ \hline
Stardog                       & \cmark & \xmark & \cmark & \xmark & \xmark & \cmark & \cmark & \mmark & \xmark & \xmark & \xmark & \cmark \\
RDFox                         & \cmark & \xmark & \xmark & \xmark & \xmark & \mmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\
query.wikidata                & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark \\
yasgui                        & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark & \cmark & \xmark & \xmark \\
Protégé                       & N/A    & N/A    & N/A    & N/A    & \cmark & N/A    & \cmark & N/A    & N/A    & N/A    & \cmark & \xmark \\
rdf-vocabularies-autocomplete & N/A    & N/A    & N/A    & N/A    & N/A    & \xmark & \xmark & \cmark & \xmark & N/A    & N/A    & \xmark \\
JSON-LD LSP                   & \cmark & \xmark & \cmark & \xmark & \xmark & \xmark & \xmark & \cmark & \xmark & \xmark & \cmark & \xmark \\
SWLS                          & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark \\
\hline
\end{tabularx}
    \caption{\label{tab:current_implementations}
    \mmark Stardog simple completion is based on a fixed list of items. 
    \mmark RDFox syntax completion is only based on predefined for SPARQL functions.
    Table listing IDE features of different Semantic Web tools.
    All eligable software supports both highlighting and completion in one form of another, 
    while no software supports formatting or renaming except for Protégé and JSONLD-LSP (our previous work). 
  }
\end{table}

\subsection{Integration of SWLS with Existing Ontology and Development Environments}

The \textit{Semantic Web Language Server} (SWLS), based on the \textit{Language Server Protocol} (LSP), is not intended as a replacement for comprehensive ontology editors such as \textit{Protégé} or general-purpose integrated development environments (IDEs) like \textit{IntelliJ IDEA} and \textit{Eclipse}.
Instead, it functions as a complementary tool that extends IDE-like capabilities to a broad range of text editors and development environments.

A key advantage of LSP-based solutions is their ability to integrate seamlessly with existing tools.
Protégé, for instance, provides a rich graphical interface for ontology engineering but also allows users to open individual ontology files in a text-based format.
When editing such files, SWLS can enhance the experience by offering syntax validation, autocompletion, and real-time diagnostics, 
thereby bridging the gap between textual and graphical ontology development.

Similarly, general-purpose IDEs such as IntelliJ IDEA and Eclipse, which support LSP, can leverage SWLS to provide semantic web-specific assistance when working with ontology files, SHACL constraints, or SPARQL queries. 
This integration allows users to remain within their familiar development environment while still benefiting from specialized semantic web tooling.

By adhering to the LSP standard, SWLS fosters interoperability between semantic web tools and mainstream development environments. 
Rather than replacing established solutions, it extends their functionality, ensuring that users can leverage the strengths of both specialized ontology tools and general-purpose development platforms.


\subsection{Comparison to Large Language Models for Semantic Web Assistance}

Recent advances in \textit{Large Language Models} (LLMs), such as \textit{ChatGPT}, have demonstrated impressive capabilities in assisting users with code generation, query formulation, and ontology-related tasks. 
However, while LLMs can serve as powerful aids in navigating the complexities of the Semantic Web, they also introduce challenges related to reliability and determinism.  

A key distinction between SWLS and LLM-based approaches lies in their underlying data sources and guarantees.
SWLS retrieves autocompletion suggestions, validation feedback, and syntax assistance directly from ontology files, ensuring that all recommendations align with formally defined schema and constraints.
This deterministic behavior is essential for correctness in ontology engineering, where adherence to established standards is crucial. 
In contrast, LLMs generate responses probabilistically based on their training data, which, while often useful, can lead to hallucinations or incorrect suggestions that do not conform to the given ontology.

Despite these limitations, LLMs could complement SWLS by assisting users in discovering relevant ontologies.
Given the vast number of existing ontologies and vocabularies, manually identifying the most appropriate resources can be challenging. 
LLMs, trained on a broad range of semantic web content, could help users locate relevant ontologies based on natural language descriptions, which SWLS could then use as a structured source of truth for deterministic editing support.  

Thus, while SWLS and LLM-based approaches serve different roles, they are not mutually exclusive. 
A hybrid approach, leveraging SWLS for deterministic operations and LLMs for exploratory tasks, could provide a more comprehensive semantic web development experience.  


